Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada

### 2025-03-01 writing/manuscript.Rmd
committing to try to recover lost appendix

---
title: "Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - bcp: BC Parks
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'freshwater-biology.csl'
# csl: 'nature.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
    - \usepackage{booktabs}
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # for Delta symbol in table
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
editor_options: 
  chunk_output_type: console
---

@@

\noindent \textbf{Appendices}: 3

@@

This manuscript leverages four main datasets: (1) a collection of GPS telemetry data; (2) historical hourly reanalyzed air temperature from the ERA5 dataset from the European Center for Medium-range Weather Forecasting; (3) resource rasters of percent forest cover, elevation, and distance from water; and (4) climate change projections under four Shared Socioeconomic Pathways [SSPs, see @riahi_shared_2017]. We detail the data sources and analyses in the sections below, and more details are available in appendices A (estimating how temperature affects movement frequency and speed when moving), B (fitting Hierarchical Resource Selection Functions), and C (simulating hourly temperature data to 2100 using the four SSPs).

@@

We obtained rasters of projected monthly average temperature at a 0.08$^\circ$ resolution in BC from 2020 to 2100 under the different SSPs via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`. Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the hourly ERA5 data for BC from 1998 to 2023 [extremes included, see @hersbach_era5_2023] to calculate within-month variance in temperature, which we defined as the variance within a given pixel, month, and year. We then modeled the estimated variance using a GAM for location and scale [GAMLS, see: @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017]. The linear predictor for the location (i.e., the mean) included smooth terms of the within-pixel monthly mean temperature (within each year), month (as a cyclic smooth), a two-dimensional smooth of space, and a tensor interaction product term of space and month. The linear predictor for the scale term (which governs the mean-variance relationship) included smooth terms of the monthly mean, month, and space. We did not include a smooth of year to avoid unrealistic projections when extrapolating beyond the range of historical data. Additional details are available in Appendix C.

@@

We simulated hourly variation in weather by assuming temperature followed a normal distribution with mean specified by the monthly `climatenaR` climate projections and variance as specified by the Gamma GAMLS. We then predicted changes in movement behavior and habitat selection as a function of the simulated temperature using the HGAMs and HRSFs. For each month within each year from 2020 to 2100, we simulated hourly weather by including temperatures from the 0.1 to the 0.9 quantile by increments of 0.1, and we weighted each quantile proportionally to the (normalized) Gaussian probability density for each quantile (Appendix C).

@@

Since `ctmm`'s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), data-informed speeds needed to be corrected so that the minimum instantaneous speed could be 0. We performed this correction by subtracting each model's mean speed while assuming speeds were $\chi^2$-distributed. The function we used is available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R). To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using the inflection point of each species' histogram of detrended speed estimates (Appendix A).

@@

\noindent In this manuscript, we separate the effects of temperature on mammalian movement into indirect and direct effects. We call "indirect" all effects that can be intervened on by altering an animal's habitat [its geographic space *sensu* @matthiopoulos_defining_2022], and we refer to all other effects as "direct". For example, an increase in temperature may push an animal to prefer higher forest cover in search of shade, and conservationists can thus mitigate these indirect effects of temperature on movement behavior by increasing the availability of forested areas, since the effect of temperature on movement is conditional on forest cover availability. In contrast, the animal may avoid moving if temperatures are above 35$^\circ$C, irrespective of its geographic and environmental space [*sensu* @matthiopoulos_defining_2022], so it is not possible to mitigate the effects of temperature on movement behavior (other than changing temperature directly). We provide Directed Acyclical Graphs in Appendix A (modeling the effects of temperature on movement rates), and Appendix B (modeling the effects of temperature on habitat selection).

@@

\noindent Species differed in mean probabilities of movement (range: 0.05 -- 0.3), mean speed when moving (range: 0.42 -- 2.7 km/day), and mean distance traveled (range: 0.04 -- 0.6 km/day), even after predicting for one-hour time intervals for all species (Table \@ref(tab:gam-intercepts)). Grizzly bears tended to move least often (probability of moving $P(M) \approx 0.05$), while wolves and pumas moved most often ($P(M) \gtrapprox 0.23$). When moving, mountain goats and southern mountain caribou moved the slowest ($\mathbb E(S|M) \approx 0.4$ m/s), while wolves had the highest mean speed when moving ($\mathbb E(S|M) \approx 2.7$ m/s). Consequently, wolves traveled, on average, 50 km/day -- 2.5 to 16.7 times further than other mammals.

@@

Including temperature in the models always increased within-sample prediction, but the change was not always substantial (deviance explained increased between 0.2% and 18.6%, with all HRSFs having an increase $\ge 4.6$), and there was relatively strong agreement between models with and without temperature (Fig. ***SX pred vs pred***), but including temperature resulted in substantially better fits for all models, especially HRSFs (all $\Delta$AIC $\le$ -342; Table \@ref(tab:temp-comparison)).

@@

\noindent The marginal effects of temperature varied in both direction and magnitude across species (Fig. ***AX***), even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; see Fig. ***AX***). The smooths' credible intervals were generally quite wide outside the typical temperature range of -20$^\circ$C to 20$^\circ$C, so the estimated trends were often not credible at the 95% level outside such range. Smooth interaction terms were well-behaved and indicated clear shifts in activity over time of day and day of year for all species. The models explained reasonably high proportions of the deviance (10.7% for the binomial model and 79.3% for the Gamma model) and had good in-sample prediction (Fig. ***AX***). All species' movement frequency and speed decreased with increasing sampling interval (Fig. ***AX***).

Most species adapted their daily and seasonal movement behavior to changes in temperature (Fig. \@ref(fig:dist)). Wolves, elk, pumas, and boreal caribou reduced their mid-day movement rates when temperatures were above 20$^\circ$C, and wolves, elk, and boreal caribou moved less on summer days above 20$^\circ$C. However, the precision of the estimated change depended on the GPS sampling frequency (Fig. ***AX***). While the `ctmm` models produced scale-independent estimates of speed [i.e., model interpretation is independent of sampling time: @noonan_scale-insensitive_2019], the accuracy, size, and stability of speed estimates still depended on the GPS sampling frequency. This is because coarsely-sampled movement data contains information on large-scale movements (e.g., range crossings, migrations) but not fine-scale movements [e.g., first-order habitat selection *sensu* @johnson_comparison_1980]. Using the boreal caribou as an example, the 13-hour sampling interval allows us to reasonably estimate the caribou's movement path at a temporal scale of approximately 13 hours (or greater), but we cannot produce reasonable movement trajectories at a much finer (e.g., hourly) scale. Consequently, we suggest being cautious when comparing estimated movement behaviors across species.

```{r dist, fig.cap="\\textbf{Temperature is a strong determinant of how far and when mammals travel.} The fill indicates the estimated effects of temperature on the relative change in distance travelled (probability of moving times mean speed when moving) over time of day (\\textbf{A}) and day of year (\\textbf{B}). Predictions in the surface plots extend to 10% of the range away from each datum. The color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to -75\\% and +300\\% (i.e., $2^{\\pm 2}$) for ease of readability."}
knitr::include_graphics('../figures/distance.png')
```

@@

Species' relative selection strength (RSS) for resources depended on temperature, and RSS was strongest for elevation and weakest for forest cover. Changes in RSS with temperature were strongest for elevation and generally weakest for distance from water, but there were no common trends across all species for any of the three resources. All species, with the exception of pumas, exhibited a temperature-dependent shift in their preference for forest cover. At higher temperatures, wolves, mountain goats, and grizzly bears became less selective for forest cover, while elk and caribou became more selective. All species shifted elevationally with temperature, although boreal caribou did not exhibit as strong of a change. As temperatures rose, elk, mountain goats, cougars, and boreal caribou increased in elevation, while boreal wolves, southern mountain caribou, and grizzly bears decreased in elevation. Most species generally remained within 5 km of water, and temperature did not affect their selection strength as much as for the other two resources. Estimated RSS values were generally most uncertain at extreme temperatures (Fig. ***BX***).

@@

\noindent The direction and magnitude of change in distance traveled varied greatly between species, but worse climate-change scenarios generally corresponded to greater absolute changes in behavior (Fig. \@ref(fig:t-dist)). Under the best-case scenario, absolute projected changes by 2100 were small (approximately 0% to 2%), while absolute projected changes by 2100 ranged from ~1% (boreal caribou) to ~25% (southern mountain caribou) under the worst-case scenario. Boreal caribou and elk showed relatively little projected changes in yearly distance traveled. Throughout BC  (Fig. \@ref(fig:s-dist)), all species but elk and boreal caribou were expected to change their movement rates by 10 to 20% in 2100, relative to 2025, but the predicted changes varied spatially due to heterogeneity in climate change throughout the province. Across all species, absolute changes were smallest under the best-case scenario and greatest under the worst-case scenario.

```{r t-dist, fig.cap="\\textbf{The direction and magnitude of changes in movement rates due to climate change varies among species, but worse climate-change scenarios often result in the greatest change and greatest uncertainty.} Lines indicate the projected relative change in mean distance travelled (probability of moving times speed when moving) due to changes in temperature within the species' current extent. Shaded areas indicate the average 95% Bayesian credible intervals for the area. Changes are relative to the predicted mean distances travelled in 2025 across the four Shared Socioeconomic Pathways (SSPs). The projections only account for changes movement behavior (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
knitr::include_graphics('../figures/distance-travelled-local-cc-predictions.png')
```

```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently.} The color scale indicates the predicted changes in distance traveled in 2100, relative to each location's average distance in 2025 across all four scenarios, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase. For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.8 and 1.25 ($\\approx 2^{\\pm 0.32}$; original data ranged 0.81 to 1.35). The predictions only account for the predicted temperature throughout the province and ignore environmental factors such as terrain slope, soil type, and forest cover."}
knitr::include_graphics('../figures/bc-displ-2100.png')
```

All species were projected to reduce selection for their current range throughout the 21^st^ century, with the exception of grizzly bears, which showed little to no change under all four the climate change scenario (\@ref:fig(t-hrsfs)). The decreases in RSS were most evident in the habitats with the top 5% RSS in the current range. Projected changes in the median-RSS habitats were relatively small, but none increased throughout the century under any of the climate change scenarios. As with distance traveled, the change in RSS in 2100 varied throughout BC for all species (Fig. \@ref(fig:s-hrsfs)). All species were predicted to exhibit a decrease in RSS throughout most of BC, with the exception of boreal caribou, which had a predicted decrease in approximately 37% of the province, including their current range. While the boreal caribou were expected to increase selection for altitudes > 1 km above sea level, the estimated selection strength was highly uncertain and not credible at the 95% level (Fig. ***SX***).

```{r, eval=FALSE}
library('dplyr')
readRDS('data/cc-rss-bc-projections-albers.rds') %>%
  group_by(lab, scenario) %>%
  summarize(p_decrease = mean(lambda < 1)) %>%
  filter(p_decrease < 0.5)
```

```{r t-hrsfs, fig.cap="\\textbf{The intensity and direction of changes in habitat selection due to climate change vary among species, but no species are expected to exhibit an increase in relative selection strength (RSS).} Lines indicate the projected change in RSS based on the 5$^{th}$ percentile (the worst 5%; bottom line), the median (the middle; bold), and 95% percentile (the best 5%; top line) habitats within the species' current extent. Changes are relative to the median RSS in 2025 across the four Shared Socioeconomic Pathways (SSPs)."}
knitr::include_graphics('../figures/rss-local-cc-predictions.png')
```

```{r s-hrsfs, fig.cap="\\textbf{Climate change will impact each species' habitat relative selection strength (RSS) differently.} The color scale indicates the predicted changes in RSS in 2100, relative to each location's average RSS in 2025 across all four scenarios, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase. For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.6}$; original data ranged 0.26 to 117)."}
knitr::include_graphics('../figures/bc-rss-2100.png')
```

@@

\noindent We demonstrated that temperature is an important determinant of how and where mammals move, even while accounting for other seasonal drivers of movement (e.g, changes in photoperiod, hormones, and resources). Disentangling temperature from other drivers allows us to predict how changes in climate will affect mammalian movement even if all other drivers remain approximately constant or adapt at negligible rates. However, as habitats warm and animals are exposed to increasingly frequent extreme and novel conditions, our ability to confidently predict mammals' responses decreases. Predicting at the fringes of the surface plots we present comes with substantial uncertainty, particularly given that the responses are nonlinear. At unusually warm temperatures, mammals (and other homeotherms) overheat, so their movement behavior and that of their competitors, predators, and/or prey can often be substantially different from the behavior at typical temperatures. As extreme temperature events become more frequent and prolonged [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], mammals will be increasingly forced to atypical behaviors, which will have the potential to substantially alter community structures and behaviors, both during such events and afterwards [@zhang_increased_2023; ]. For instance, changes in climate and phenology impact the life history and behavior of many hibernating mammals [@wells_life_2022], but hot weather also affects mammals' sleep quality [e.g., wild boars, *Sus scrofa*: @mortlock_sleep_2024] and likelihood to enter torpor [Australian eastern long-eared bat, *Nyctophilus bifax*: @fjelldal_nightly_2021]. These changes can alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024]. At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase their ability to find food or be predated upon [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021] as well as the overall ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023]. However, these changes are likely to have complex consequences for population and ecosystem structures and dynamics as prey, predators, and prey experience altered and novel seasonal cycles. especially given the increasingly common climate "weirding" [@bunnell_global_2011], which includes not only warmer temperatures and more frequent and intense heat waves but also an increase in various other extreme events [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], such as cold spells, floods and droughts, plant and animal die-offs and sudden growth surges, and other changes in population dynamics and behaviors [sometimes referred to as "black swan" events -- see: @logares_black_2012; @anderson_black-swan_2017].

>>>>

---
title: "Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - bcp: BC Parks
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'freshwater-biology.csl'
# csl: 'nature.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
    - \usepackage{booktabs} # to prevent tables from splitting
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # for Delta symbol in table
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
editor_options: 
  chunk_output_type: console
---

@@

\noindent \textbf{Appendices}: 1

@@

This manuscript leverages four main datasets: (1) a collection of GPS telemetry data; (2) historical hourly reanalyzed air temperature from the ERA5 dataset from the European Center for Medium-range Weather Forecasting; (3) resource rasters of percent forest cover, elevation, and distance from water; and (4) climate change projections under four Shared Socioeconomic Pathways [SSPs, see @riahi_shared_2017]. We detail the data sources and analyses in the sections below.<!--, and more details are available in appendices A (estimating how temperature affects movement frequency and speed when moving), B (fitting Hierarchical Resource Selection Functions), and C (simulating hourly temperature data to 2100 using the four SSPs).-->

@@

We obtained rasters of projected monthly average temperature at a 0.08$^\circ$ resolution in BC from 2020 to 2100 under the different SSPs via the `climatenaR` package [version 1.0, @burnett_climatenar_2023] for `R`. Since the climate projections only provided monthly means and ranges but no measures of variance or distributional assumptions, we used the hourly ERA5 data for BC from 1998 to 2023 [extremes included, see @hersbach_era5_2023] to calculate within-month variance in temperature, which we defined as the variance within a given pixel, month, and year. We then modeled the estimated variance using a GAM for location and scale [GAMLS, see: @rigby_generalized_2005; @stasinopoulos_generalized_2007; section 7.9 in @wood_generalized_2017]. The linear predictor for the location (i.e., the mean) included smooth terms of the within-pixel monthly mean temperature (within each year), month (as a cyclic smooth), a two-dimensional smooth of space, and a tensor interaction product term of space and month. The linear predictor for the scale term (which governs the mean-variance relationship) included smooth terms of the monthly mean, month, and space. We did not include a smooth of year to avoid unrealistic projections when extrapolating beyond the range of historical data.<!-- Additional details are available in Appendix C.-->

@@

We simulated hourly variation in weather by assuming temperature followed a normal distribution with mean specified by the monthly `climatenaR` climate projections and variance as specified by the Gamma GAMLS. We then predicted changes in movement behavior and habitat selection as a function of the simulated temperature using the HGAMs and HRSFs. For each month within each year from 2020 to 2100, we simulated hourly weather by including temperatures from the 0.1 to the 0.9 quantile by increments of 0.1, and we weighted each quantile proportionally to the (normalized) Gaussian probability density for each quantile<!-- (Appendix C)-->.

@@

Since `ctmm`'s movement models assume stochastic but non-zero speeds (i.e., a single, stochastic moving state), data-informed speeds needed to be corrected so that the minimum instantaneous speed could be 0. We performed this correction by subtracting each model's mean speed while assuming speeds were $\chi^2$-distributed. The function we used is available on GitHub at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/functions/detrend_speeds.R). To avoid artifacts due to excessively small, non-zero speeds, we determined whether an animal was moving or not using the inflection point of each species' histogram of detrended speed estimates, as $k$-nearest neighbors did not discriminate between states well (Fig. S3).

@@

\noindent In this manuscript, we separate the effects of temperature on mammalian movement into indirect and direct effects. We call "indirect" all effects that can be intervened on by altering an animal's habitat [its geographic space *sensu* @matthiopoulos_defining_2022], and we refer to all other effects as "direct". For example, an increase in temperature may push an animal to prefer higher forest cover in search of shade, and conservationists can thus mitigate these indirect effects of temperature on movement behavior by increasing the availability of forested areas, since the effect of temperature on movement is conditional on forest cover availability. In contrast, the animal may avoid moving if temperatures are above 35$^\circ$C, irrespective of its geographic and environmental space [*sensu* @matthiopoulos_defining_2022], so it is not possible to mitigate the effects of temperature on movement behavior (other than changing temperature directly). We provide the relevant Directed Acyclical Graphs in the Appendix (Figs. S1-S2).

@@

\noindent Data were relatively scarce for temperatures lower than $-20^\circ$C or above $20^\circ$C (Fig. S4). Species differed in overall mean probabilities of movement (range: 0.05 -- 0.3), mean speed when moving (range: 0.42 -- 2.7 km/day), and mean distance traveled (range: 0.04 -- 0.6 km/day), even after predicting for one-hour time intervals for all species (Table \@ref(tab:gam-intercepts)). Grizzly bears tended to move least often (probability of moving $P(M) \approx 0.05$), while wolves and pumas moved most often ($P(M) \gtrapprox 0.23$). When moving, mountain goats and southern mountain caribou moved the slowest ($\mathbb E(S|M) \approx 0.4$ m/s), while wolves had the highest mean speed when moving ($\mathbb E(S|M) \approx 2.7$ m/s). Consequently, wolves traveled, on average, 50 km/day -- 2.5 to 16.7 times further than other mammals.

@@

Including temperature in the models always increased within-sample prediction, but the change was not always substantial (deviance explained increased between 0.2% and 18.6%, with all HRSFs having an increase $\ge 4.6$), and there was relatively strong agreement between models with and without temperature (Fig. S5), but including temperature resulted in substantially better fits for all models, especially HRSFs (all $\Delta$AIC $\le$ -342; Table \@ref(tab:temp-comparison)).

@@

\noindent The effects of temperature on movement rates varied in both direction and magnitude across species, even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; see Figs. S4-S6). <!--The smooths' credible intervals were generally quite wide outside the typical temperature range of -20$^\circ$C to 20$^\circ$C, so the estimated trends were often not credible at the 95% level outside such range. -->Smooth interaction terms were well-behaved and indicated clear shifts in activity over time of day and day of year for all species. The models explained reasonably high proportions of the deviance (10.7% for the binomial model and 79.3% for the Gamma model) and had good in-sample prediction (Fig. S9). All species adapted their daily and seasonal movement behavior to changes in temperature (Fig. \@ref(fig:dist)). Wolves, elk, pumas, and boreal caribou reduced their mid-day and summer movement rates when temperatures were above 20$^\circ$C. However, uncertainties around the estimated effects were higher at extreme temperatures due to data scarcity.

All species' movement frequency and speed decreased with increasing sampling interval (Fig. S10). While the `ctmm` models produced scale-independent estimates of speed [i.e., model interpretation is independent of sampling time: @noonan_scale-insensitive_2019], the accuracy, size, and stability of speed estimates still depended on the GPS sampling frequency. This is because coarsely-sampled movement data contains information on large-scale movements (e.g., range crossings, migrations) but not fine-scale movements [e.g., first-order habitat selection *sensu* @johnson_comparison_1980]. Using the boreal caribou as an example, the 13-hour sampling interval allows us to reasonably estimate the caribou's movement path at a temporal scale of approximately 13 hours (or greater), but we cannot produce reasonable movement trajectories at a much finer (e.g., hourly) scale. Consequently, we suggest being cautious when comparing estimated movement behaviors across species even though all predictions have been corrected to the hourly timescale by predicting for 1-hour time intervals (i.e., `dt_hours = 1`).

```{r dist, fig.cap="\\textbf{Temperature is a strong determinant of how far and when mammals travel.} The fill indicates the estimated effects of temperature on the relative change in distance travelled (probability of moving times mean speed when moving) over time of day on June $1^{st}$ (\\textbf{A}) and day of year at 12:0 PDT (\\textbf{B}). Predictions in the surface plots extend to 10% of the range away from each datum. The color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to -75\\% and +300\\% (i.e., $2^{\\pm 2}$) for ease of readability."}
# not specifying dt, or doy for A and time for B because change is relative
knitr::include_graphics('../figures/distance.png')
```

@@

Species' relative selection strength (RSS) for resources depended on temperature, and RSS was strongest for elevation and weakest for forest cover. Changes in RSS with temperature were strongest for elevation and generally weakest for distance from water, but there were no common trends across all species for any of the three resources. All species, with the exception of pumas, exhibited a temperature-dependent shift in their preference for forest cover. At higher temperatures, wolves, mountain goats, and grizzly bears became less selective for forest cover, while elk and caribou became more selective. All species shifted elevationally with temperature, although boreal caribou did not exhibit as strong of a change. As temperatures rose, elk, mountain goats, cougars, and boreal caribou increased in elevation, while boreal wolves, southern mountain caribou, and grizzly bears decreased in elevation. Most species generally remained within 5 km of water, and temperature did not affect their selection strength as much as for the other two resources. Estimated RSS values were generally most uncertain at extreme temperatures (Fig. S11).

@@

\noindent The direction and magnitude of change in distance traveled varied greatly between species, but worse climate-change scenarios generally corresponded to greater absolute changes in behavior (Fig. \@ref(fig:t-dist)). Under the best-case scenario, absolute projected changes by 2100 were small (approximately 0% to 2%), while under the worst-case scenario changes ranged from ~1% (boreal caribou) to ~25% (southern mountain caribou). Elk and boreal caribou showed relatively little projected changes in yearly distance traveled. Throughout BC (Fig. \@ref(fig:s-dist)), all species but elk and boreal caribou were expected to change their movement rates by 10 to 20% in 2100, relative to 2025, but the predicted changes varied spatially due to heterogeneity in climate change throughout the province. Across all species, absolute changes were smallest under the best-case scenario and greatest under the worst-case scenario.

```{r t-dist, fig.cap="\\textbf{The direction and magnitude of changes in movement rates due to climate change varies among species, but worse climate-change scenarios often result in the greatest change and greatest uncertainty.} Lines indicate the projected relative change in mean distance travelled (probability of moving times speed when moving) due to changes in temperature within the species' current extent. Shaded areas indicate the 0.025 and 0.975 quantiles of the estimated means. Changes are relative to the predicted mean distances travelled in 2025 across the four Shared Socioeconomic Pathways (SSPs). The projections only account for changes movement behavior (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
knitr::include_graphics('../figures/distance-travelled-local-cc-predictions.png')
```

```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently.} The color scale indicates the predicted changes in distance traveled in 2100, relative to each location's average distance in 2025 across all four scenarios, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase. For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.8 and 1.25 ($\\approx 2^{\\pm 0.58}$; original data ranged 0.72 to 1.29). The predictions only account for the predicted temperature throughout the province and ignore environmental factors such as terrain slope, soil type, and forest cover."}
knitr::include_graphics('../figures/bc-displ-2100.png')
```

All species exhibited a decrease in RSS within their current range, irrespective of climate change scenario, but changes were often were only between -1% and -3% under the best-case scenario. Under the worst-case scenario, changes in RSS ranged from -8% to -14% (\@ref:fig(t-hrsfs)). As with distance traveled, the change in RSS in 2100 varied throughout BC for all species (Fig. \@ref(fig:s-hrsfs)). All species were predicted to exhibit a decrease in RSS throughout most of BC, with the exception of boreal caribou, which had a predicted decrease in approximately 37% of the province, including their current range. However, this was in great part due to the estimated increase in selection for high-elevation areas that are well outside the population's current range.

```{r, eval=FALSE}
# find species that decrease in RSS in less than half of BC
library('dplyr')
readRDS('data/cc-rss-bc-projections-albers.rds') %>%
  group_by(lab, scenario) %>%
  summarize(p_decrease = mean(lambda < 1)) %>%
  filter(p_decrease < 0.5)
```

```{r t-hrsfs, fig.cap="\\textbf{All species are expected to exhibit a decrease in relative selection strength (RSS) for their current range, irrespective of climate change scenario, but decreases are stronger in the worse-case scenario.} Lines indicate the projected change in median RSS within the species' current extent. Changes are relative to the median RSS in 2025 across the four Shared Socioeconomic Pathways (SSPs)."}
knitr::include_graphics('../figures/rss-local-cc-predictions.png')
```

```{r s-hrsfs, fig.cap="\\textbf{Climate change will impact each species' habitat relative selection strength (RSS) differently.} The color scale indicates the predicted changes in RSS in 2100, relative to each location's average RSS in 2025 across all four scenarios, such that 1 indicates no change, values < 1 indicate a decrease, and values > 1 indicate an increase. For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.6}$; original data ranged 0.26 to 117). The increase in boreal caribou's RSS for high elevations at warmer temperatures was highly uncertain and often not credible at the 95\\% level."}
knitr::include_graphics('../figures/bc-rss-2100.png')
```

@@

\noindent We demonstrated that temperature is an important determinant of how and where mammals move, even while accounting for other seasonal drivers of movement (e.g, changes in photoperiod, hormones, and resources). Disentangling temperature from other drivers allows us to predict how changes in climate will affect mammalian movement even if all other drivers remain approximately constant or adapt at negligible rates. However, as habitats warm and animals are exposed to increasingly frequent extreme and novel conditions, our ability to confidently predict mammals' responses decreases. Predicting at the fringes of the surface plots we present comes with substantial uncertainty, particularly given that the responses are nonlinear. At unusually warm temperatures, mammals (and other homeotherms) overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behavior and that of their competitors, predators, and/or prey can often be substantially different from the behavior at typical temperatures. As extreme temperature events become more frequent and prolonged [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], mammals will be increasingly forced to atypical behaviors, which will have the potential to substantially alter community structures and behaviors, both during such events and afterwards [@zhang_increased_2023]. For instance, changes in climate and phenology impact the life history and behavior of many hibernating mammals [@wells_life_2022], and hot weather can also affects mammals' sleep quality [e.g., wild boars, *Sus scrofa*: @mortlock_sleep_2024] and likelihood to enter torpor [e.g., Australian eastern long-eared bat, *Nyctophilus bifax*: @fjelldal_nightly_2021]. Our results show that warmer temperatures cause grizzly bears to more further earlier in the year (Fig. \@ref(fig:dist)), which may alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024]. At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase their ability to find food or be predated upon [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021] as well as the overall ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023]. These changes will likely have complex consequences for population and ecosystem structures and dynamics as prey, predators, and competitors experience altered seasonal cycles and increasingly common climate "weirding" [@bunnell_global_2011], including warmer temperatures and more frequent and intense extreme events [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], such as heat waves, cold spells, floods and droughts, plant and animal die-offs, and sudden growth surges [sometimes referred to as "black swan" events -- see: @logares_black_2012; @anderson_black-swan_2017].

***continue reading from HERE***
